Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools

### 2024-03-26 paper.qmd
Update table code, fix #40

However, travel to school and work account for less than half (39%) of all trips in Ireland, excluding 'returning home' trips, according to the 2022 National Household Travel Survey (NHTS) [@national2022].
This represents a substantial decline in the proportion of trips covered by POWSCAR data compared with pre-COVID survey results, with 51% of non-return trips being for work or education according to the 2017 release of the same report [@national2017].
To tackle this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips.
An overview of the trip purposes used in CRUSE is presented in @tbl-trip_purposes.

@@

```{r}
#| label: tbl-trip_purposes
# For PDF:
knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.", booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
  kableExtra::column_spec(1, width = "13em") |>
  kableExtra::column_spec(2, width = "15em") |>
  kableExtra::column_spec(3, width = "12em") |>
  kableExtra::column_spec(4, width = "5em")
# For HTML:
# knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.")
```

>>>>

However, travel to school and work account for less than half (39%) of all trips in Ireland, excluding 'returning home' trips, according to the 2022 National Household Travel Survey (NHTS) [@national2022].
This represents a substantial decline in the proportion of trips covered by POWSCAR data compared with pre-COVID survey results, with 51% of non-return trips being for work or education according to the 2017 release of the same report [@national2017].
To tackle this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips.
An overview of the trip purposes used in CRUSE is presented in @tbl-trip-purposes.

@@

```{r}
#| label: tbl-trip-purposes
# For PDF:
knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.", booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
  kableExtra::column_spec(1, width = "13em") |>
  kableExtra::column_spec(2, width = "15em") |>
  kableExtra::column_spec(3, width = "12em") |>
  kableExtra::column_spec(4, width = "5em")
# For HTML:
# knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.")
```

### 2024-03-26 paper.tex
Update table code, fix #40

However, travel to school and work account for less than half (39\%) of
all trips in Ireland, excluding `returning home' trips, according to the
2022 National Household Travel Survey (NHTS) \citep{national2022}. This
represents a substantial decline in the proportion of trips covered by
POWSCAR data compared with pre-COVID survey results, with 51\% of
non-return trips being for work or education according to the 2017
release of the same report \citep{national2017}. To tackle this issue,
we developed a spatial interaction modeling methodology to estimate the
number of trips between each OD pair for additional trip purposes. The
classification of trip purposes used in the CRUSE Tool was guided mainly
by the trip purpose classification found within the National Household
Travel Survey (NHTS), but with the addition of categories based on the
comprehensive POWSCAR data, and the need to include recreational trips
and multi-stage trips. An overview of the trip purposes used in CRUSE is
presented in Figure~\ref{fig-trip-purposes}.

\includegraphics{images/paste-4.png}

\caption{\label{fig-trip-purposes}Everyday trip purposes used in the
CRUSE Tool.}

@@

@@

\includegraphics{images/mermaid-figure-1.png}

@@

Thoughout the project, we undertook a series of workshops and interviews
with stakeholders in Ireland to gather feedback on the approach and
resulting tool. The feedback was generally positive, with stakeholders
commenting on the need for the tool, lack of data on cycling potential
proving a barrier to decision-making and investment, and the potential
for the tool. At an end of project workshop in December 2023,
stakeholders were asked to provide feedback on the tool and its
potential use in practice. Quotes and suggestions from these workshops,
and a subsequent interview with a Senior Transport Planning using the
evidence in their everyday work in Transport Infrastructure Ireland, and
presented in this section (with quotes being attributed to named
individuals where permission has been granted).

Early in the project, we presented the approach to practitioners in
Kildare and Limmerick. The feedback from the workshop in Kildare
highlighted the need for the tool for practitioners at the county level,
as highlighted in the following quote:

\begin{quote}
``It's a missing piece of evidence that will help new projects get
built'' (Senior Engineer, Sustainable Transport \& Traffic Management,
Kildare County Council)
\end{quote}

Practitioners in Kildare County Council that the evidence generated
\emph{was already valuable} the development of new cycle routes,
especially when making the case during planning applications. They
emphasized that cycle networks are often fragmented, and the tool allows
them to demonstrate how individual local schemes can contribute to a
larger, interconnected network. An unexpected piece of feedback from the
workshop was the preference for visualizations that directly compare the
fast and quiet routes, as they recognize the importance of both types of
routes. In response to this we provided the `Route types' page for each
county, although this functionality is not available in the landing page
map due to other feedback requesting a simpler interface. In comparison
with another tool (the National Transport Authority's cycling propensity
tool), the CRUSE tool was seen by practitioners in Kildare as more
user-friendly, with the ability to generate evidence at the route level,
which is crucial for planning new cycling infrastructure. Overall, the
workshop participants concluded that the approach filled gap in evidence
and will greatly assist in the development of new cycling projects.

@@

In a subsequent interview with a practitioner using the evidence
generated by the CRUSE approach in their everyday work, we received the
following feedback. The outputs were a ``great tool for getting started
with project outline documents'' and ``ideal for generating a strategic
rationale for estimating demand and building the business case in the
absence of cycle count data'' (Declan Keenan, Senior Transport Planner
working on active travel, personal communication). Declan also noted the
alignment between local knowledge, cycle count data, and the outputs of
the approach: ``It's a good representation of what's going on'' and the
networks allign well in terms of network ``shape'' and the proportion of
trips of trips on different parts of the network.

Findings from the qualitative results also suggest changes which have
not been implemented due to time constraints, including the provision of
route netework data showing only ``gold standard'' POWSCAR data (we did
make this data available in the data downloads page), the need for
further validation of results (especially in rural areas where counter
datasets are scarce) and the need for training on the tool to ensure
that it is well-understood and used effectively. These and other
potential areas for further research and development are discussed in
the next section.

>>>>

However, travel to school and work account for less than half (39\%) of
all trips in Ireland, excluding `returning home' trips, according to the
2022 National Household Travel Survey (NHTS) \citep{national2022}. This
represents a substantial decline in the proportion of trips covered by
POWSCAR data compared with pre-COVID survey results, with 51\% of
non-return trips being for work or education according to the 2017
release of the same report \citep{national2017}. To tackle this issue,
we developed a spatial interaction modeling methodology to estimate the
number of trips between each OD pair for additional trip purposes. The
classification of trip purposes used in the CRUSE Tool was guided mainly
by the trip purpose classification found within the National Household
Travel Survey (NHTS), but with the addition of categories based on the
comprehensive POWSCAR data, and the need to include recreational trips
and multi-stage trips. An overview of the trip purposes used in CRUSE is
presented in Table~\ref{tbl-trip-purposes}.

\caption{\label{tbl-trip-purposes}Trip purposes used in the CRUSE Tool.}

\\
\toprule
Trip purpose & Description & Primary source(s) & Confidence\\
\midrule
\cellcolor{gray!10}{Work (commute)} & \cellcolor{gray!10}{Commuting to/from workplaces} & \cellcolor{gray!10}{Census 2016 POWSCAR Data} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
Primary Education & Primary school trips & Census 2016 POWSCAR Data & Dept. of Education Schools database,High confidence - data both origins and destinations for trip purposes\\
\cellcolor{gray!10}{Secondary Education} & \cellcolor{gray!10}{Secondary school trips} & \cellcolor{gray!10}{Census 2016 POWSCAR Data} & \cellcolor{gray!10}{Dept. of Education Schools database,High confidence - data both origins and destinations for trip purposes}\\
Tertiary Education & Tertiary education trips & Census 2016 POWSCAR Data - Geodirectory categories for 'Tertiary Education' and 'Other Education' & High confidence - data both origins and destinations for trip purposes\\
\cellcolor{gray!10}{Social} & \cellcolor{gray!10}{NHTS classification, including trips to/from sports facilities, leisure, and cultural destinations (e.g. restaurants, cinemas, gyms etc.)} & \cellcolor{gray!10}{Trip rates from NHTS} & \cellcolor{gray!10}{Geodirectory businesses classified according to relevant NACE codes,"Medium Confidence - Method should identify clusters of activity for this trip purpose, although individual trips or destinations come with significant uncertainty"}\\
\addlinespace
Shopping & NHTS classification, including trips to/from supermarkets and shops & Trip rates from NHTS & Geodirectory businesses classified according to relevant NACE codes,"Medium Confidence - Method should identify clusters of activity for this trip purpose, although individual trips or destinations come with significant uncertainty"\\
\cellcolor{gray!10}{Personal/other} & \cellcolor{gray!10}{Combination of two NHTS classifications, and includes trips to medical, personal services and others} & \cellcolor{gray!10}{Trip rates from NHTS} & \cellcolor{gray!10}{Geodirectory businesses classified according to relevant NACE codes,"Medium Confidence - Method should identify clusters of activity for this trip purpose, although individual trips or destinations come with significant uncertainty"}\\
Tourism/recreational & Trips by non-resident visitors to/from visitor attractions & Geodirectory listings for accommodation (origins) and destinations (destinations) & Failte Ireland accommodation database, Failte Ireland attraction database,Low confidence - lack of data regarding tourist trip rates\\
\cellcolor{gray!10}{Public Transport} & \cellcolor{gray!10}{Trips to/from public transport} & \cellcolor{gray!10}{NTA Heavy Rail Census} & \cellcolor{gray!10}{NA}\\
\bottomrule

@@

\includegraphics[width=3.31in,height=\textheight]{images/fig5.png}

@@

\includegraphics{images/fig5.png}

@@

Throughout the project, we undertook a series of workshops and
interviews with stakeholders in Ireland to gather feedback on the
approach and resulting tool. The feedback was generally positive, with
stakeholders commenting on the need for the tool, lack of data on
cycling potential proving a barrier to decision-making and investment,
and the potential for the tool. At an end of project workshop in
December 2023, stakeholders were asked to provide feedback on the tool
and its potential use in practice. Quotes and suggestions from these
workshops, and a subsequent interview with a Senior Transport Planning
using the evidence in their everyday work in Transport Infrastructure
Ireland, and presented in this section (with quotes being attributed to
named individuals where permission has been granted).

Early in the project, we presented the approach to practitioners in
Kildare and Limerick. The feedback from the workshop in Kildare
highlighted the need for the tool for practitioners at the county level,
as highlighted in the following quote:

\begin{quote}
``It's a missing piece of evidence that will help new projects get
built'' (Dónal Hodgins, Senior Engineer, Sustainable Transport \&
Traffic Management, Kildare County Council)
\end{quote}

Practitioners in Kildare County Council found that the evidence
generated \emph{was already valuable} for the development of new cycle
routes, especially when making the case during planning applications.
They emphasized that cycle networks are often fragmented, and the tool
allows them to demonstrate how individual local schemes can contribute
to a larger, interconnected network. An unexpected piece of feedback
from the workshop was the preference for visualizations that directly
compare the fast and quiet routes, as they recognize the importance of
both types of routes. In response to this we provided the `Route types'
page for each county, although this functionality is not available in
the landing page map due to other feedback requesting a simpler
interface. In comparison with another tool for transport planning, CRUSE
was seen by practitioners in Kildare as more user-friendly. They found
the evidence at the route level (unavailable in other tools) crucial for
planning new cycling infrastructure. Overall, the workshop participants
concluded that the approach filled gap in evidence and will greatly
assist in the development of new cycling projects.

@@

In a subsequent interview with a practitioner using the evidence
generated by the CRUSE approach in their everyday work, we received the
following feedback. The outputs were found to be a ``useful tool for
getting schemes started; and provides good detail to supplement the
strategic rationale for intervention within Project Outline Documents,
via the cycle friendliness maps and the baseline / near market cyclist
estimates''. As a result, the CRUSE tool ``is ideal for generating a
strategic rationale for estimating demand and building the business case
in the absence of cycle count data; it is also very useful at the
detailed design stage and the CRUSE estimates can (and have been used)
to supplement the economic appraisal of cycling schemes'' (Declan
Keenan, Senior Transport Planner TII working on active travel, personal
communication). Declan also noted a strong alignment in the South Dublin
area between cycle count data, and the outputs of the approach: ``It's
forms a strong representation of what's going on in the areas where TII
have collated some cycling count data, e.g.~in the vicinity of the M50
in the areas of Ballinteer and Sandyford. Some additional network
checking against count data would also be useful, to ensure good
representation across other settlements.'' Where cycling data is
available, the networks align well in terms of network ``shape'' and the
proportion of trips on different parts of the network.

There were two suggestions from Declan which have not yet been
implemented: the provision of data showing POWSCAR and recreational
estimates/data separately, and the need for training on the tool to
ensure that all staff can use it effectively. These, and other potential
improvements, are discussed in the next section.
